---
title: "FBref"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r}
library(tidyverse)
library(ggrepel)
library(tidymodels)
library(embed)
library(here)
```

```{r}
raw_df <- read_csv(here("data", "fb-data_2020-21_players_share.csv"))
```

```{r}
df <- raw_df %>% 
  #too many positions, reduce
  mutate(simplified_position = case_when(
        position %in% c('DF,MF', 'MF,DF') ~ 'DM',
        position %in% c('DF,FW', 'FW,DF') ~ 'M',
        position %in% c('MF,FW', 'FW,MF') ~ 'AM',
        position == 'DF' ~ 'D',
        position == 'MF' ~ 'M',
        position == 'FW' ~ 'F',
        position == 'GK' ~ 'G',
        position == 'GK,MF' ~ 'G',
        TRUE ~ position
      )
    ) %>% 
  filter(player != "Salvador Ferrer")
```

```{r}
head(df)
```

```{r}




versus_plot <- function(x_parameter, y_parameter){
  df %>%
  filter(mins> 900) %>% 
  ggplot(aes({{x_parameter}}, {{y_parameter}} )) +
  geom_point(aes(fill = primary_color, colour =  secondary_color), shape = 21) +
  geom_text_repel(aes(label = short_name)) +
  scale_fill_identity(labels = raw_df$primary_color,
                      breaks = raw_df$primary_color) +
  scale_color_identity(labels = raw_df$secondary_color,
                      breaks = raw_df$secondary_color) +
  theme_light() +
  labs(caption = "Source: FBref.com",
       subtitle = "Players with > 900 minutes in the 2020/2021 season across the 'top five' leagues") +
  theme(plot.title = element_text(face="bold"))
}


  
  
```

```{r}
versus_plot(shots_p_90, shots_on_target_p_90) +
  labs(title = "Shots vs shots on target (per 90)",
       x = "Shots",
       y = "Shots on target")
```

```{r}
versus_plot(attempted_dribble_p_90, success_dribble_p_90) +
  labs(title = "Attempted vs successful dribbles (per 90)",
       x = "Attempted dribbles",
       y = "successful dribbles")
```

```{r}
versus_plot(pass_attempted_p_90, pass_cmp_p_90) +
  labs(title = "Attempted vs completed passses (per 90)",
       x = "Attempted passes",
       y = "Completed passes") +
  #remove a player with 0,0 for some reason
  xlim(10,120)
```

```{r}
versus_plot(tackle_attempted_p_90, tackle_won_p_90) +
  labs(title = "Attempted vs won tackles (per 90)",
       x = "Attempted tackles",
       y = "Tackles won") 
```

```{r}
versus_plot(npxG_p_90,npG_p_90) +
  labs(title = "Expected vs actual penalty pen goals (per 90)",
       x = "Expected non penalty goals",
       y = "Actual non penalty goals") +
 geom_abline()
```

# Unsupervised Learning

```{r}
unsupervised_df <- df %>% 
  filter(mins> 900) %>% 
  select(player:league, simplified_position, contains("_p_90"))



```

## PCA

### Create PCA recipe

Normalise and prep recipe. Provide an ID for later skree plot

```{r}


pca_recipe <- recipe(~., data=unsupervised_df)%>%
  step_normalize(all_numeric()) %>%
  step_pca(all_numeric(), id = "pca") %>% 
  prep() 
```

PCA plot

```{r}
pca_recipe %>% 
  juice() %>% 
  ggplot(aes(PC1, PC2, colour= simplified_position)) +
  geom_point() +
  geom_text_repel(aes(label = short_name)) +
  theme_light() +
  labs(caption = "Source: FBref.com",
       subtitle = "Players with > 900 minutes in the 2020/2021 season across the 'top five' leagues",
       title = "PCA analysis") +
  theme(plot.title = element_text(face="bold"))
  
```

Scree plot

```{r}
pca_recipe %>% 
  broom::tidy(type = "variance", id = "pca") %>% 
  filter(component <= 10) %>% 
  pivot_wider(names_from = terms, values_from = value) %>% 
  janitor::clean_names()  %>% 
  ggplot() +
  geom_col(aes(component, percent_variance)) +
  geom_line(aes(component, cumulative_percent_variance)) +
  
  theme_light() +
  theme(plot.title = element_text(face="bold")) +
  labs(title = "PCA component importance",
       x = "Component",
       y = "Variance")
```

### Find similar players

"Juice" the data

```{r}
comps <- pca_recipe %>% 
  juice()
```

Get the percent variance each PC is responsible for

```{r}
importance <- pca_recipe %>% 
  broom::tidy(type = "variance", id = "pca") %>% 
  filter(component <= 5) %>% 
  filter(terms == "percent variance") %>% 
  pivot_wider(names_from = component, values_from = value) %>% 
  clean_names() %>% 
  select(x1:x5)
```

Scale each PC by the variance it is responsible for. Find the distance between each row

```{r}

# combind the PC and vairance impact
combined <- bind_cols(comps, importance)

#scale
impact <- combined[,3:7] * combined[,8:12]

#distance matrix
distances <- as.matrix(dist(impact, method = "euclidean"))

#name each column
colnames(distances) <- unsupervised_df$player


#convert to tibble
comparison <- as_tibble(distances) 
```

Recombine this matrix with the starting dataset

```{r}
matrix_data <- unsupervised_df %>% 
  select(1:7) %>% 
  bind_cols(comparison)

```

Find similar players

```{r}

matrix_data %>% 
  select(1:7, player_of_interest = `Houssem Aouar`) %>% 
  arrange(player_of_interest) 
```

# UMAP

## Create UMAP recipe

```{r}
umap_recipe <- recipe(~., data=unsupervised_df)%>%
  step_normalize(all_numeric()) %>%
  step_umap(all_numeric(), id = "umap") %>% 
  prep() 
```

```{r}
umap_recipe %>% 
  juice() %>% 
  ggplot(aes(umap_1, umap_2, colour= simplified_position)) +
  geom_point() +
  geom_text_repel(aes(label = short_name)) +
  theme_light() +
  labs(caption = "Source: FBref.com",
       subtitle = "Players with > 900 minutes in the 2020/2021 season across the 'top five' leagues",
       title = "UMAP analysis") +
  theme(plot.title = element_text(face="bold"))
```

## Find similar players

```{r}
impact <- umap_recipe %>% 
  juice() %>% 
  select(contains("umap"))


#distance matrix
distances <- as.matrix(dist(impact, method = "euclidean"))

#name each column
colnames(distances) <- unsupervised_df$player


#convert to tibble
comparison <- as_tibble(distances) 
```

```{r}
matrix_data <- unsupervised_df %>% 
  select(1:7) %>% 
  bind_cols(comparison)
```

```{r}
matrix_data %>% 
  select(1:7, player_of_interest = `David de Gea`) %>% 
  arrange(player_of_interest) 
```
